Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-02-17 articleSummaries.qmd
Updated Intro Take 2
Brad edits 2/16

-   What is the goal of the paper?

-   Why is it important?

-   How is it solved? - methods

-   Results/limitations, if any.

>>>>

[@Ming-Chiang2014]

https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=8e45bb12b4bf296f002977caf11e969d5cc90a02

This article uses Cox Proportional Hazards model to predict bankruptcies in companies in the Taiwanese stock market. The model used uses 12 variables from companies found between 2003 and 2009 to classify companies into the categories of distressed vs non-distressed to indicate if they are at risk of bankruptcy. Each variable had reasoning and/or sources to explain the reasoning behind using them in the model along with statistical justification. The resulting model had an accuracy of 88%.

[@Bustan2018]

https://iopscience.iop.org/article/10.1088/1742-6596/1028/1/012230/pdf

This article describes a Cox Proportional Hazards Model to predict the probability of survival in breast cancer patients. The model uses six variables to make predictions and provides an explanation of the risk factors that are the most heavily weighted when deciding a patient's outlook. The article includes a table showing the survival probabilities of patients based on the values of their corresponding covariates. The article concludes the most important factor in surviving breast cancer is to catch it in as early of a stage as possible and recommends more outreach programs to encourage individuals to get themselves examined by a doctor regularly.

[@Kuitunen2021]

This article explores the proportionality assumption of cox proportional hazard models on arthroplasty research datasets. It accounts for over 1100 studies. The article concludes that the proportionality assumption cannot be ignored, as without it the results of the models created are unreliable. It is critical for the proportional hazards assumption to be tested properly, or inaccurate predictions will be made,

[@Smith2003]

In this article, Cox Proportional Hazard modeling was used for modeling time until event while accounting for other variables. The author explains how Cox PH models work mathematically and the strenths of Cox PH models. Cox PH models are able to accomodate incomplete or "censored" data which is prevalent in survival data. In many clinical trials, a patient death may not be observed because the trial period will end not allowing enough time for the event to occur. Health datasets were analyzed with hospitalization and demographic data to predict hospitalization time after inhaling oil well fire smoke. The model creates a probability for each individual's chances of being hospitalized.

[@Abeysekera2009]

This article summarizes the use of Shoenfeld's global test for testing assumptions of Cox PH models. It mentions that other methods for checking assumptions have an aspect of subjectivity to them, but this method shines because it does not. The proportional hazards assumption is the most important aspect of a dataset when trying to analyze data with a Cox PH model, and the Schoenfeld's global test is an effective method to test the assumption. The article includes visualizations and ways to interpret the visuals. The example provided in the paper shows the conclusion of which covariates were shown to be in violation of the assumption in a better way than typical methods. The article recommends Shoenfeld's global test as the ideal method for testing assumptions.

[@Seung2023]

https://www.elifecycle.org/archive/view_article?pid=lc-3-0-8

This article compares Kaplan-Meier survival analysis with Cox Proportional Hazards models. It summarizes how effective and necessary the models are in the health industry, and how important it is to pass the proportional hazards assumption to ensure an accurate model. It demonstrates the reason Cox models are so widely used over Kaplan-Meier because they're able to accommodate multiple covariates into predictions. The paper also explains how to handle datasets that break the assumptions of Cox PH models. It also goes over the time dependent regression models that can be used when the assumptions are broken. The author concludes that each model mentioned has its merits, but Cox PH models are the most popular models because of their flexibility and ability to include multiple covariates.

### 2025-02-17 index.qmd
Brad edits 2/16

---
title: "Predicting Remaining Useful Life of NASA Aircraft Engines "
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

@@

>>>>

---
title: "Predicting Remaining Useful Life of NASA Aircraft Engines"
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

@@

## Introduction (take 3)

As the COVID-19 pandemic demonstrated, the ability to predict the
survival of a patient based on their symptoms and medical history is a
major benefit as society deals with health crises such as pandemics. Those
predictions would assist medical practitioners in triage and allow
researchers the ability to generate predictions on the individuals who
are at the greatest risk of losing their lives due to an illness. While
the medical world is the most obvious use of survival predictions, the
same methods can be applied to the predicting bank failure, the
remaining useful life of machines, machinery maintenance timelines, and
insurance likelihood of payouts. There are various survival methods that
can be examined, including Cox proportional hazards model which will be
examined in this paper.

Cox proportional hazards (CPH) models are statistical regression models specializing in
modeling time-to-event predictions with survival data. Survival data is data with a value
for time and an event. For example, in a study investigating the survival likelihood of
mechanical equipment time would begin when the equipment begins functioning and would
conclude in the event the equipment stops functioning. [@Smith2003] CPH models can use
single or multiple covariates as a means of creating predictions until failure and are
capable of dealing with survival data that contains censored data as well.[@Seung2023]
Censored data is when the information about an individual in a study is only known for a
certain period of time, such as the length of the study [@klein2005, 1]. The models also
have the advantage of being semi-parametric which allows more flexibility as compared to
other models. [@Ming-Chiang2014]

An additional limitation of CPH comes from the data in the covariates
selected and how can they can become biased. Two studies done using UK
health data demonstrated how bias can impact the values for the
covariates when relying on patient submitted information. The first was
a study on the connection between salt intake and the development of
anxiety or depression after 14.5 years [@wang2025] and the second was a
study on the association between postpartum depression and the
development of two or more chronic diseases [@zhang2025]. Both of those
studies relied on patient submitted health data and survey responses
that were supplemented with death records and other accessible health
records thus creating a possible bias in their data which could
negatively influence their model as their covariates may not have
accurately reflected the true data. These articles provide examples of
the care that needs to be taken when working with CPH models.

Continuing with using CPH outside of health data, we will study the
remaining useful life of a NASA aircraft engine using data provided by
NASA for the study on propagation modeling [@saxena2008]. From the
literature review, we have noted that we will need to be concerned with
the constant hazards ratio within our data and determine how we will
accommodate censored data when creating our model.
